The Gamma: Data Exploration through Iterative Prompting

### 2020-09-12 wip-design/paper.tex
Edit overview, move implementation

\documentclass[manuscript,review,anonymous]{acmart}
\usepackage{hyperref}

@@

% \usepackage{xspace}
% \usepackage{fancyvrb}
% \usepackage{inconsolata}
% \usepackage{xcolor}
% \usepackage{tikz}
% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
%     \draw (0,0) circle (1);
%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     \end{tikzpicture}}
% \newcommand*{\priorityc}[1]{
% \resizebox{0.8em}{0.8em}{
%   {\protect\tikz{ \protect
%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     } }}}
%
% \definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
% \DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
% \newcommand{\kvd}[1]{\textbf{#1}}
% \newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
% \newcommand{\tg}{The Gamma\xspace}
% \usepackage{balance}       % to better equalize the last page
% \usepackage{graphics}      % for EPS, load graphicx instead
% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
% \usepackage{txfonts}
% \usepackage{mathptmx}
% \usepackage[pdflang={en-US},pdftex]{hyperref}
% \usepackage{color}
% \usepackage{booktabs}
% \usepackage{textcomp}

@@

\subsubsection*{Iterative prompting}
The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
valid data exploration scripts can be constructed by repeatedly choosing from a list
of offered options. This way, non-programmers can write entire scripts through auto-complete,
without learning a programming language, but they still produce transparent and reproducible
source code. In other words, iterative prompting turns auto-complete from a programmer assistance
tool into a non-expert programming mechanism. A crucial feature is that iterative prompting
only offer operations that are valid in a given context and that it offer all such operations;
it is both correct and complete.

\subsubsection*{Date exploration}
The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
The user accesses data available in a structured format. They make several experiments to find an
interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
may choose to view the results as a table or a basic chart before publishing their analysis. The
Gamma makes such simple programming with data simple enough for non-experts. Scraping and cleaning
of messy data or building rich data visualizations is outside of the scope of this paper, but
exposing those using the iterative prompting approach is an interesting future challenge.

\subsubsection*{Contributions}
The Gamma is available (non-anonymously) at \href{http://thegamma.net}{\textcolor{blue}{thegamma.net}},
both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
evaluate the design principles behind the project:

@@

\section{Related work}

@@

\subsubsection*{Code completion for data science}

@@

\subsubsection*{Notebooks and business intelligence tools}

@@

\subsubsection*{Easier programming tools}
We aim to build an easy to use and learn programming system. Many approaches to this
goal have been tried. Victor \cite{learnable,principle} introduced design principles that
inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
immediate feedback to help programmers understand how code relates to output and
exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.
A system combining textual language with visualization also exists for graph querying \cite{guess}.
To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
the interactions with code more natural. The Gamma is live in that our editor gives an instant
preview of the results.

@@

\subsubsection*{Programming without writing code}
There are two main approaches to programming where
the user does not write code. In programming by example \cite{byexample}, the user gives
examples of desired results. This has been used, e.g.~for specifying data transformations
in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
In direct manipulation \cite{direct}, a program is specified by directly interacting with the
output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
modification in this context. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

\subsubsection*{Guestures and data entry}
Although our focus is on program construction, our work can be positioned in the
broader context of input methods. Akin to Dasher \cite{dasher}, our system provides a way of
navigating through a complete space of options, while on-screen feedforward \cite{octopocus} allows
efficient selection in guesture-based interfaces. Those provide compelling alternatives to
auto-completion menus, although the efficiency of input methods is not an issue in programming.

\section{Overview}

design motivated by exploratory question - what if we could program with autocomplete
no talking to journalists hence shorter paper

maybe talk about "design space"

\newpage

@@

\tg is a text-based programming environment that allows non-experts to create simple data exploration
scripts using a single interaction principle -- choosing an item from an auto-complete list.
It supports a range of data sources including tabular data, graph data and data cubes.

\subsection{Querying Travel Expenses}
To introduce \tg, we walk through a simple problem that a local journalist might want to solve.
The UK government publishes travel expense claims by members of the House of Lords. A journalist
wants to find out which of the members representing the Kent county spend the most on travel.
The following shows a subset of the data:\footnote{ \url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }

\begin{thegamma}
\textbf{Name, County, Days Attended, Days Away, Travel Costs}
Lord Adonis, London, 8, 0, 504
Baroness Afshar, Yorkshire, 2, 0, 0
Lord Alderdice, Oxfordshire, 3, 0, 114
Lord Alli, London, 5, 0, 0
Baroness Amos, London, 3, 0, 0
\end{thegamma}

After the analyst imports the CSV file (through a web interface), the environment is initialized
with code that refers to the imported variable \ikvd{expenses}. The analyst then types `.' (dot):

\begin{thegamma}
expenses.
\end{thegamma}
\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords1}

The type provider for tabular data allows analysts to construct simple queries. It first offers
a list of operations that the analyst might want to perform such as grouping, filtering and sorting.
To find members of the House of Lords from Kent, the analyst chooses \ikvd{filter data},
types `.' and then chooses \ikvd{County is} from the offered list, types `.' and starts
typing Kent:

\begin{thegamma}
expenses
  .'filter data'.'County is'.Ke
\end{thegamma}
\vspace{-0.4em}\qquad\includegraphics[width=0.5\columnwidth]{figures/lords2}

The completion list is generated from the values in the \ikvd{County} column of the dataset.
After selecting \ikvd{Kent}, the live preview is updated to only show records according to the
specified filter:

\includegraphics[width=1\columnwidth]{figures/lords3}

To finish specifying filtering conditions, the analyst chooses \ikvd{then} and is offered the
same list of querying operations as in the first step. To sort House of Lords members by their
travel costs, she now chooses \ikvd{sort data} and types `.' (dot):

\begin{thegamma}
expenses
  .'filter data'.'County is'.Kent.then
  .'sort data'.
\end{thegamma}
\qquad\includegraphics[width=0.5\columnwidth]{figures/lords4}

The auto-complete offers a list of columns that can be used for sorting, each with ascending
(default) and descending order option. After choosing one or more sort keys, the analyst selects
the \ikvd{then} member and is, again, offered the list of querying operations. They use
\ikvd{paging} to get top 5 records and \ikvd{get series} to obtain a data series with just
the House of Lords member name and their travel expenses.

\begin{thegamma}
expenses
  .'filter data'.'County is'.Kent.then
  .'sort data'.'by Travel Costs descending'.then
  .paging.take(5)
  .'get series'.'with key Name'.'and value Travel Costs'
\end{thegamma}

When the code evaluates to a data series with a categorical (textual) key and a numerical value,
\tg switches from displaying the result as a table to a column chart:

\includegraphics[width=1\columnwidth]{figures/lords5}
\vspace{-1.5em}

\subsection{Querying via Iterative Prompting}
In the example, the journalists constructs a query that filters and sorts a data table. This is
not unlike writing a query in the SQL language. However, the whole query is constructed through
a single interaction mechanism that we refer to as \emph{iterative prompting}. The user triggers
the interaction by typing `.' at the end of the query constructed so far. She is then offered a
list of available operations in the current context. This may include a choice of top-level
operations (e.g.~at the start and then again after choosing \ikvd{then}) or possible
parameters for an operation (such as possible sorting keys and directions after choosing
\ikvd{sort data}). The user then merely needs to choose one of the available options, rather than
learning the syntax and semantics of SQL in order to know what constructs can be used in the
given context.
We discuss the important aspects of The Gamma design in the next section and then discuss the
integration of specific data sources in detail, revisiting the type provider for querying
data tables.

@@

\subsection{Making Complex Things Possible}
Although the focus on The Gamma is simple data exploration and almost all the use cases we
envision can be achieved using the iterative prompting interaction principle, there remain a
few cases where more flexibility is needed. We aim to \emph{make simple things easy and
complex things possible}. The language thus supports a small number of additional constructs that
can only be used through text editing. The following example illustrates three additional
constructs:

\begin{thegamma}
\kvd{let} topTravel =
  expenses
    .'sort data'.'by Travel Costs descending'.then
    .paging.take(5).'get series'
    .'with key Name'.'and value Travel Costs'

charts.column(topTravel)
  .setTitle("House of Lords members by travel expenses")
  .setColors(["red","blue","green"])
\end{thegamma}

First, The Gamma allows operations with parameters such as \ikvd{take(3)} or \ikvd{setTitle("...")}.
This is currently needed when writing a query that skips or takes first N elements from a table.
The remaining features are not needed for basic data exploration, but allow other scenarios, such
as manual creation of charts. The \ikvd{let} construct can be used to define (immutable) variables
and The Gamma also supports lists written as \ikvd{[1,2,3]}.

@@

\section{Implementation}
\label{sec:implementation}
The Gamma consists of a programming language with a web-based coding environment and a number
of type providers that provide access to various kinds of data sources. The system is implemented
in F\# and compiled to JavaScript using Fable.

\subsection{The Gamma Language}
In The Gamma, a program is a sequence of commands. A command can be either a variable declaration
or an expression that evaluates to a value such as a data table or a chart.
An expression is a reference to a data source followed by a chain of member accesses.
A member can be either an ordinary member or an operation which takes a list of parameters
enclosed in parentheses. When the member name contains non-alphanumerical characters such as
whitespcae, it needs to be written in quotes. Our implementation supports a few other features,
but those are not used in this paper.

The Gamma uses a type system to infer what members are available at a given point in a chain.
A type is an object with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by a type providers for individual data sources.
The types are generated lazily and are only computed when a member appears in code.

The web-based coding environment uses the Monaco\footnote{\url{https://microsoft.github.io/monaco-editor/}}
editor and implemts auto-completion feature that is triggered when the user types `.' and relies
on information from the type system. The can display the value of a current command inline
as in Figure~\ref{fig:thegamma}, or values of all commands on the side as in Figure~\ref{fig:sidebyside}.
In the latter case, it can also combine one or more data series into a single chart.

\begin{figure}
\centering
\includegraphics[scale=0.28]{figures/cubetp}
\caption{Exploring data cubes. The World Bank provider allows users to first choose
  a country or a year and then an indicator.}
\label{fig:cubetp}
\end{figure}

\subsection{Data Cube Type Provider}
Our first type provider allows users explore data form a data cube. For example, the World
Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
countries in the world each year. The data set is a three-dimensional cube with dimensions
corresponding to countries, indicators and years. The following example, also shown in
Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:

\subsection{Tabular Data Type Provider}

\begin{figure}
\centering
\includegraphics[scale=0.28]{figures/graphtp}
\vspace{0.5em}
\caption{Exploring graph databases. The user specifies a path through the data, possibly with
  placeholders to select multiple nodes.}
\label{fig:graphtp}
\end{figure}

\subsection{Graph Database Type Provider}
Our third type provider allows users explore data from graph databases. A graph database
consists of nodes representing entities and relationships between them. In the context of
data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.

>>>>

\documentclass[manuscript,review,anonymous]{acmart}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{hyperref}

@@

% \usepackage{xspace}
\usepackage{fancyvrb}
% \usepackage{inconsolata}
\usepackage{xcolor}
% \usepackage{tikz}
% \newcommand*{\priority}[1]{\begin{tikzpicture}[scale=0.12]%
%     \draw (0,0) circle (1);
%     \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     \end{tikzpicture}}
% \newcommand*{\priorityc}[1]{
% \resizebox{0.8em}{0.8em}{
%   {\protect\tikz{ \protect
%     \draw[line width=1mm, black] (0,0) circle (1); \fill[fill opacity=1,fill=black] (0,0) -- (90:1) arc (90:90-#1*3.6:1) -- cycle;
%     } }}}
%
\definecolor{kvdclr}{rgb}{0.3,0.0,0.8}
\DefineVerbatimEnvironment{thegamma}{Verbatim}{fontfamily=zi4,numbers=left,xleftmargin=6mm,fontsize=\small,commandchars=\\\{\}}
% %\newcommand{\kvd}[1]{\textcolor{kvdclr}{#1}}
\newcommand{\kvd}[1]{\textbf{#1}}
\newcommand{\ikvd}[1]{{\fontfamily{zi4}\selectfont\small #1}}
% \usepackage{balance}       % to better equalize the last page
% \usepackage{graphics}      % for EPS, load graphicx instead
% \usepackage[T1]{fontenc}   % for umlauts and other diaeresis
% \usepackage{txfonts}
% \usepackage{mathptmx}
% \usepackage[pdflang={en-US},pdftex]{hyperref}
% \usepackage{color}
% \usepackage{booktabs}
% \usepackage{textcomp}

@@

\subsubsection*{Iterative Prompting}
The key idea in The Gamma, which we term the \emph{iterative prompting} principle, is that all
valid data exploration scripts can be constructed by repeatedly choosing from a list
of offered options. This way, non-programmers can write entire scripts through auto-complete,
without learning a programming language, but they still produce transparent and reproducible
source code. In other words, iterative prompting turns auto-complete from a programmer assistance
tool into a non-expert programming mechanism. A crucial feature is that iterative prompting
only offer operations that are valid in a given context and that it offer all such operations;
it is both correct and complete.

\subsubsection*{Data Exploration}
The Gamma focuses on data exploration of the kind illustrated in Figure~\ref{fig:thegamma}.
The user accesses data available in a structured format. They make several experiments to find an
interesting way of looking at the data, e.g.~by applying different aggregations or filters. They
may choose to view the results as a table or a basic chart before publishing their analysis. The
Gamma makes such simple programming with data simple enough for non-experts. Scraping and cleaning
of messy data or building rich data visualizations is outside of the scope of this paper, but
exposing those using the iterative prompting approach is an interesting future challenge.

\subsubsection*{Contributions}
The Gamma is available (non-anonymously) at \url{http://thegamma.net},
both as a JavaScript library and a hosted data exploration service. In this paper, we describe and
evaluate the design principles behind the project:

@@

\section{Related Work}

@@

\subsubsection*{Code Completion for Data Science}

@@

\subsubsection*{Notebooks and Business Intelligence Tools}

@@

\subsubsection*{Easier Programming Tools}
We aim to build an easy to use and learn programming system. Many approaches to this
goal have been tried. Victor \cite{learnable,principle} introduced design principles that
inspired many to build live programming systems \cite{review,liveroad,lighttable} that give
immediate feedback to help programmers understand how code relates to output and
exploratory systems \cite{variolite,exploratory} that assist with completing open-ended tasks.
A system combining textual language with visualization also exists for graph querying \cite{guess}.
To avoid difficulties with editing code as text, some systems use structured editors~\cite{structure-based,livenut,lamdu}.
In Subtext \cite{subtext,directprog} the language itself is co-designed with the editor to make
the interactions with code more natural. The Gamma is live in that our editor gives an instant
preview of the results.

@@

\subsubsection*{Programming without Writing Code}
There are two main approaches to programming where
the user does not write code. In programming by example \cite{byexample}, the user gives
examples of desired results. This has been used, e.g.~for specifying data transformations
in spreadsheets and data extraction \cite{spreadsheetpbe,flashextract}.
In direct manipulation \cite{direct}, a program is specified by directly interacting with the
output. This has been used in the visual domain \cite{sketchnsketch}, but also for data querying
\cite{dynamicq,vlang}. The VQE language~\cite{visage} also considers how to allow code reuse and
modification in this context. Direct manipulation can also support data exploration by letting
users partially edit queries,~e.g. by changing quantifiers as in DataPlay~\cite{dataplay}.

\subsubsection*{Guestures and Data Entry}
Although our focus is on program construction, our work can be positioned in the
broader context of input methods. Akin to Dasher \cite{dasher}, our system provides a way of
navigating through a complete space of options, while on-screen feedforward \cite{octopocus} allows
efficient selection in guesture-based interfaces. Those provide compelling alternatives to
auto-completion menus, although the efficiency of input methods is not an issue in programming.

\begin{figure}[b]
\includegraphics[width=1\columnwidth]{figures/thegamma-walk}
\caption{Constructing a script that charts the top 5 members of the House of Lords for Kent, based
on their travel costs.}
\label{fig:walkthrough}
\end{figure}

The review of related work suggests that there is an unexplored point in the design space of tools
for data exploration. Although various efforts make text-based programming easier, e.g. by providing
high-level declarative abstractions, most systems that target non-experts shy away from code, by using
either programming by example, natural language or graphical user interface. As spelled out in
Section~\ref{sec:design}, text-based programming has notable benefits for public-facing data analyses
such as learnability and transparency. Our work is thus motivated primarily by the question whether we
can make text-based programming easy enough for non-experts? The study presented in Section~\ref{sec:study}
shows that this is, indeed, possible at least for typical data querying tasks.
The Gamma consists of a programming language, a web-based coding environment and a number of type providers
that enable access to various kinds of data sources. It allows non-experts to create programs using the
\emph{iterative prompting} interaction principle -- by repeately selecting an item from an auto-complete list.
We start with a walkthrough of the system (Section~\ref{sec:overview-walk}), before looking at details of
The Gamma language (Section~\ref{sec:overview-lang}) and individual type providers (Section~\ref{sec:overview-tps}).

\subsection{Querying Travel Expenses}
\label{sec:overview-walk}

To introduce The Gamma, we walk through a simple problem that a local journalist might want to solve.
The UK government publishes travel expense claims by members of the House of Lords. We want to find
out which of the members representing the Kent county spend the most on travel.
The following shows a subset of the data:\footnote{Full data set has been obtained from
\url{https://www.parliament.uk/mps-lords-and-offices/members-allowances/house-of-lords/holallowances/} }

\begin{thegamma}
\textbf{Name, County, Days Attended, Days Away, Travel Costs}
Lord Adonis, London, 8, 0, 504
Baroness Afshar, Yorkshire, 2, 0, 0
Lord Alderdice, Oxfordshire, 3, 0, 114
Lord Alli, London, 5, 0, 0
\end{thegamma}
% Baroness Amos, London, 3, 0, 0

\noindent
The data is available as a CSV file. After the analyst imports the file through a web interface,
the environment is initialized with code that refers to the imported data as \ikvd{expenses}
and she starts exploring the data using the type provider for tabular data (Section~\ref{sec:overview-tps}),
following the steps illustrated in Figure~\ref{fig:walkthrough}:

\begin{enumerate}
\item The analyst types `.' (dot) to trigger auto-completion on \ikvd{expenses}. The type provider
  offers a list of operations that the analyst might want to perform such as grouping,
  filtering and sorting.

\item To find the House of Lords with the largest spending, the analyst chooses the
  \ikvd{sort data} operation. Next, she is offered a list of possible arguments based on the
  schema of the tabular data and chooses the desired one.

\item The Gamma evaluates the source code on-the-fly and shows a preview
  of results. After choosing the sorting key, the analyst sees a table with House of Lords members
  from more remote counties of the UK.

\item To finish specifying the (possibly compound) sorting key, the analyst chooses \ikvd{then}
  and is offered the same list of querying operations as in the first step. To obtain House
  members from Kent, she chooses \ikvd{filter data}. To navigate through the offered list more
  efficiently, she types first two characters of the name.

\item After selecting \ikvd{County is} to specify the desired type of condition, the analyst types
  `.' and is offered a list of options based on the values of the \ikvd{County} column in the
  source data set. She types \ikvd{Ke} and selects \ikvd{Kent}.

\item The analyst chooses \ikvd{then} and is,
  again, offered the list of querying operation. She uses \ikvd{paging} to get top 5 records,
  which requires typing \ikvd{5} as the argument. She then uses the \ikvd{get series} operation
  to obtain a data series associating travel expenses with a name, which is automatically
  visualized using a bar chart.
\end{enumerate}

\noindent
The constructed code is not unlike an SQL query, except that the whole script is constructed using
iterative prompting, by repeatedly selecting one of the offered members. Those represent both
operations, such as \ikvd{sort by} and arguments, such as \ikvd{Kent}. The only exception
is when the analyst needs to type the number \ikvd{5} to specify the number of items to take.

\subsection{The Gamma Programming Environment}
\label{sec:overview-lang}

The Gamma consists of a text-based programming language with a web-based coding environment,
based on the Monaco editor\footnote{For more information, see \url{https://microsoft.github.io/monaco-editor/}},
and type providers that provide access to data tables, graph databases and data cubes.

\subsubsection*{The Gamma Language}
A program in The Gamma is a sequence of commands. A command can be either a variable declaration
or an expression that evaluates to a value such as a data table or a chart.
An expression is a reference to a data source followed by a chain of member accesses.
A member can be either an ordinary member such as \ikvd{paging} or an operation which takes a
list of parameters enclosed in parentheses as in \ikvd{take(5)}. When the member name contains
non-alphanumerical characters it is  written in quotes such as \ikvd{\textquotesingle sort by\textquotesingle}.

The Gamma uses a type system to infer what members are available at a given point in a chain.
Each expression has a type with a list of members that, in turn, have their own types.
The types are not built-in, but are generated by type providers for individual data sources.
The programming environment for The Gamma is based on a text editor. When the user types `.'
the editor triggers auto-completion and retrives a list of available members based on the type
information. The programming environment evaluates scripts on-the-fly and shows a preview as
illustrated in Figure~\ref{fig:thegamma}.

\subsubsection*{Making Complex Things Possible}
As illustrated by the \ikvd{take(5)} operation, there is a handful of situations where The
Gamma does not yet fully support the iterative prompting principle. The language supports
a small number of other features that can be used by more advanced users through text editing:

\vspace{0.3em}
\begin{thegamma}
\kvd{let} topTravel = expenses.'sort data'.'by Travel Costs descending'.then
  .paging.take(5).'get series'.'with key Name'.'and value Travel Costs'
charts.column(topTravel).setColors(["red","blue","green"])
  .setTitle("House of Lords members by travel expenses")
\end{thegamma}
\vspace{0.3em}

\noindent
First, The Gamma allows operations with parameters such as \ikvd{take(5)} or \ikvd{setTitle("...")}.
This is currently needed when writing a query that skips or takes first N elements from a table.
The remaining features are not needed for basic data exploration. The \ikvd{let} construct can be
used to define (immutable) variables and The Gamma also supports lists written as \ikvd{[1,2,3]}.
Advanced language features are currently used when building custom charts, but we expect that a
charting library compatible with iterative prompting would alleviate the need for most of those.

\subsection{Providers}
\label{sec:overview-tps}

all the important logic

define DSLs

\subsubsection*{Data Cube Type Provider}
Our first type provider allows users explore data form a data cube. For example, the World
Bank\footnote{\url{https://data.worldbank.org/}} collects a range of indicators about many
countries in the world each year. The data set is a three-dimensional cube with dimensions
corresponding to countries, indicators and years. The following example, also shown in
Figure~\ref{fig:sidebyside}, uses the provider to access CO2 emission data for United States:

\subsubsection*{Tabular Data Type Provider}

\begin{figure}
\centering
\begin{subfigure}[b]{0.5\textwidth}
  \centering
  \includegraphics[scale=0.25]{figures/cubetp}
  \vspace{0.5em}
  \caption{The World Bank provider allows users to first choose
    a country or a year and then an indicator.}
  \label{fig:cubetp}
\end{subfigure}
\hfill
\begin{subfigure}[b]{0.45\textwidth}
  \centering
  \includegraphics[scale=0.28]{figures/graphtp}
  \caption{The user specifies a path through the data, possibly with
    placeholders to select multiple nodes.}
  \label{fig:graphtp}
\end{subfigure}
\vspace{-0.5em}
\caption{Design of type providers for exploring data cubes and graph databases.}
\label{fig:three graphs}
\end{figure}

\subsubsection*{Graph Database Type Provider}
Our third type provider allows users explore data from graph databases. A graph database
consists of nodes representing entities and relationships between them. In the context of
data journalism, graph databases have been used for example in The Panama Papers reporting \cite{panama}.

\newpage
% ==================================================================================================

@@

@@

@@
